Follow-up Samples

### 2017-07-21 submission/manuscript.tex
complete revision edits incorporation

\section{The effect of treatment on the microbiota of patients diagnosed
with colonic
lesions}\label{the-effect-of-treatment-on-the-microbiota-of-patients-diagnosed-with-colonic-lesions}

@@

\textbf{Results.} There were large changes to the bacterial communities
associated with treatment across the three groups. The communities from
patients with carcinomas changed significantly more than those with
adenoma following treatment (P-value \textless{} 0.001). There was no
significant change in the microbiota between patients with adenoma and
advanced adenoma, or between patients with advanced adenoma and
carcinoma (P-value \textgreater{} 0.05). Although treatment was
associated with intrapersonal changes, the change in the abundance of
individual OTUs to treatment was not consistent within diagnosis groups
(P-value \textgreater{} 0.05). Because the distribution of OTUs across
patients and diagnosis groups was irregular, we used the Random Forest
machine learning algorithm to identify groups of OTUs that allowed us to
successfully distinguish between pre and post-treatment samples for each
of the diagnosis groups. Although the three models successfully
differentiated between the pre and post-treatment samples, there was
little overlap between the OTUs that were indicative of treatment. Next,
we used a larger cohort that contained individuals with normal colons
and those with adenomas, advanced adenomas, and carcinomas to determine
whether individuals who underwent treatment were more likely to have
OTUs associated with normal colons. We again built Random Forest models
and measured the change in the positive probability of having one of the
three diagnoses. Only patients who had carcinomas experienced a
significant decrease in positive probability of having a lesion after
treatment (P-value \textless{} 0.05), indicating that the microbial
milieu of the colon more closely resembled that of a normal colon.
Finally, we tested whether the type of treatment impacted the microbiota
of those diagnosed with carcinomas and were unable to detect any
significant differences in characteristics of these communities between
individuals treated with surgery alone and those treated with
chemotherapy or chemotherapy and radiation (P-value \textgreater{}
0.05).

\textbf{Conclusions.} By better understanding the response of the
microbiota to treatment for adenomas and carcinomas, it is likely that
biomarkers will be validated that can be used to quantify the risk of
recurrence and the likelihood of survival.

@@

\textbf{\emph{Treatment alters the bacterial community structure of
patients diagnosed with colonic lesions.}} Within our 67-person cohort
we tested whether the microbiota of patients with adenoma (N = 22),
advanced adenoma (N = 19), or carcinoma (N = 26) had any broad
differences between pre and post-treatment samples {[}Table 1{]}. The
structure of the microbial communities of the pre and post-treatment
samples differed, as measured by the \(\theta\)\textsubscript{YC} beta
diversity metric {[}Figure 1A{]}. We found that the communities obtained
pre and post-treament among the patients with carcinomas changed
significantly more than those patients with adenoma (P-value \textless{}
0.001). There were no significant differences in the amount of change
observed between the patients with adenoma and advanced adenoma or
between the patients with advanced adenoma and carcinoma (P-value
\textgreater{} 0.05). Next, we tested whether there was a consistent
direction in the change in the community structure between the pre and
post-treatment samples for each of the diagnosis groups {[}Figure
1B-D{]}. We only observed a consistent shift in community structure for
the patients with carcinoma when using a PERMANOVA test (adenoma P-value
= 0.999, advanced adenoma P-value = 0.945, and carcinoma P-value =
0.005). Finally, we measured the number of observed OTUs, Shannon
evenness, and Shannon diversity in the pre and post-treatment samples
and did not observe a significant change for any of the diagnosis groups
(P-value \textgreater{} 0.05) {[}Table S1{]}.

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups (P-value
\textgreater{} 0.05). It is likely that high inter-individual variation
and the irregular distribution of OTUs across individuals limited the
statistical power of the test. To overcome these problems we developed
Random Forest models to identify collections of OTUs that would allow us
to differentiate between pre and post-treatment samples from each of the
diagnosis groups. To limit the likelihood that the models would overfit
the data because of the relatively small number of subjects in each
group, we restricted our models to only incorporate 10 OTUs. Despite
this restriction, the models performed well (adenoma AUC range = 0.69 -
0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma AUC range =
0.82 - 0.98). Interestingly, the 10 OTUs that were used for each model
had little overlap with each other {[}Figure 2{]}. These results support
the earlier community-wide analysis where we observed that the treatment
had an impact on the overall community structure; however, the effect of
treatment was not consistent across patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma more
closely resemble those of a normal colon.}} Next, we determined whether
treatment changed the microbiota in a way that the post-treatment
communities resembled that of patients with normal colons. To test this,
we used an expanded cohort of 423 individuals that were diagnosed under
the same protocol as having normal colons or colons with adenoma,
advanced adenoma, or carcinoma {[}Table 2{]}. We then constructed Random
Forest models to classify the study samples, with the 3 diagnosis groups
(adenoma, advanced adenoma, or carcinoma), or having a normal colon. The
models performed well (adenoma AUC range =0.62 - 0.72, advanced adenoma
AUC range = 0.68 - 0.77, carcinoma AUC range = 0.84 - 0.90; Figure S1).
The OTUs that were incorporated into the adenoma and advanced adenoma
models largely overlapped and those OTUs that were used to classify the
carcinoma samples were largely distinct from those of the other two
models {[}Figure 3A{]}. Among the OTUs that were shared across the three
models were those populations generally considered beneficial to their
host (e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
\emph{Roseburia}) {[}Figures 3B{]}. Although many of these OTUs were
also included in the model differentiating between patients with normal
colons and those with carcinoma, this model also included OTUs
affiliated with populations that have previously been associated with
carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
\emph{Parvimonas}) {[}24--26{]} {[}Figure S2{]} with some individuals
showing are marked decrease in relative abundance {[}Figure S3{]}.
Finally, we applied these three models to the pre and post-treatment
samples for each diagnosis group and quantified the change in the
positive probability of the model. A decrease in the positive
probability would indicate that the microbiota more closely resembled
that of a patient with a normal colon. There was no significant change
in the positive probability for the adenoma or advanced adenoma groups
{[}Figure 4{]}. The positive probability for the pre and post-treatment
samples from patients diagnosed with carcinoma significantly decreased
with treatment, suggesting a shift toward a normal microbiota for most
individuals. Only, 6 of the 26 patients (23.08\%) who were diagnosed
with a carcinoma had a higher positive probability after treatment; one
of those was re-diagnosed with carcinoma on the follow up visit. These
results indicate that, although there were changes in the microbiota
associated with treatment, those experienced by patients with carcinoma
after treatment yielded gut bacterial communities of greater similarity
to that of a normal colon.

\textbf{\emph{Difficult to identify effects of specific treatments on
the change in the microbiota.}} The type of treatment that the patients
received varied across diagnosis groups. Those with adenomas and
advanced adenomas received surgical resection (adenoma, N=4; advanced
adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=18;
advanced adenoma, N=15) and those with carcinomas received surgical
resection (N=12), surgical resection with chemotherapy (N=9), and
surgical resection with chemotherapy and radiation (N=5). We focused on
the patients with carcinoma and pooled those patients that received
chemotherapy with those that received chemotherapy and radiation to
improve our statistical power. We did not observe a significant
difference in the effect of these treatments on the number of observed
OTUs, Shannon diversity, or Shannon evenness (P-value \textgreater{}
0.05). Furthermore, there was not a significant difference in the effect
of the treatments on the amount of change in the community structure
(P-value = 0.298). Finally, the change in the positive probability was
not significantly different between the two treatment groups (P-value =
0.999). Due to the relatively small number of samples in each treatment
group, it was difficult to make a definitive statement regarding the
specific type of treatment on the amount of change in the structure of
the microbiota.

@@

Our study focused on comparing the microbiota of patients diagnosed with
adenoma, advanced adenoma, and carcinoma before and after treatment. For
all three groups of patients, we observed changes in their microbiota.
After treatment, the microbiota of patients with carcinoma changed
significantly more than the other groups. This change resulted in
communities that more closely resembled those of patients with a normal
colon. This may suggest that treatment for carcinoma is not only
successful for removing the carcinoma but also at reducing the
associated bacterial communities. Understanding the effect of treatment
on the microbiota of those diagnosed with carcinomas may have important
implications for reducing disease recurrence. It is intriguing that it
may be possible to use microbiome-based biomarkers to not only predict
the presence of lesions but to assess the risk of recurrence.

Patients diagnosed with adenoma and advanced adenoma, however, did not
experience a shift towards a community structure that resembled those
with normal colons. This may be due to the fundamental differences
between the features of adenomas and advanced adenomas and carcinoma.
Specifically, carcinomas may create an inflammatory milieu that would
impact the structure of the community and removal of that stimulus would
alter said structure. It is possible that the difference between the
microbiota of patients with adenoma and advanced adenoma and those with
normal colons is subtle. This is supported by the reduced ability of our
models to correctly classify patients with adenomas and advanced
adenomas relative to those diagnosed with carcinomas {[}Figure S1{]}.
Given the irregular distribution of microbiota across patients in the
different diagnosis groups, it is possible that we lacked the
statistical power to adequately characterize the change in the
communities following treatment.

@@

This study expands upon existing research that has established a role
for the microbiota in tumorigenesis and that demonstrated the utility of
microbiome-based biomarkers to predict the presence of colonic lesions.
The most exciting future direction from the current study is the
possibility that markers within the microbiota could be used to evaluate
the effect of treatment and predict recurrence for those diagnosed with
carcinoma. If such an approach is effective, it might be possible to
target the microbiota as part of adjuvant therapy. Our data provides
additional evidence on the importance of the microbiota in tumorigenesis
by addressing the recovery of the microbiota after treatment and opens
interesting avenues of research into how these changes may affect
recurrence.

@@

\textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
have been previously reported in Baxter, et al {[}24{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between one and two weeks after initial colonoscopy. The bacterial
community has been shown to normalize back to a pre-colonoscopy
community within this time period {[}27{]}. Our study cohort consisted
of 67 individuals with an initial sample as described and a follow up
sample obtained between 188 - 546 days after treatment of lesion
{[}Table 1{]}. Patients were diagnosed by colonoscopic examination and
histopathological review of any biopsies taken. Patients were classified
as having advanced adenoma if they had an adenoma greater than 1 cm,
more than three adenomas of any size, or an adenoma with villous
histology. This study was approved by the University of Michigan
Institutional Review Board. All study participants provided informed
consent and the study itself conformed to the guidelines set out by the
Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
described by Kozich, et al. {[}28{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between four sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing.}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}28{]}. The general workflow using mothur
included merging paired-end reads into contigs, filtering for low
quality contigs, aligning to the SILVA database {[}29{]}, screening for
chimeras using UCHIME {[}30{]}, classifying with a naive Bayesian
classifier using the Ribosomal Database Project (RDP){[}31{]}, and
clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm
{[}32{]}. The number of sequences for each sample was rarefied to 10523
to minimize the impacts of uneven sampling.

\textbf{\emph{Model Building.}} The Random Forest {[}33{]} algorithm was
used to create the three models used to classify pre and post-treatment
samples by diagnosis (adenoma, advanced adenoma, or carcinoma). The
total number of individuals in the pre versus post-treatment models was
67 individuals. There were a total of 22 individuals in the pre versus
post-treatment adenoma model, 19 individuals in the pre versus
post-treatment advanced adenoma model, and 26 individuals in the pre
versus post-treatment carcinoma model {[}Table 1{]}.

Similarily, the Random Forest {[}33{]} algorithm was also used to create
the three models used to classify normal versus diagnosis. The total
number of individuals in the normal versus diagnosis models was 423
individuals {[}Table 2{]}. There were a total of 239 individuals in the
normal versus adenoma model, 262 individuals in the normal versus
advanced adenoma model, and 266 individuals in the normal versus
carcinoma model {[}Table 2{]}.

All models included only OTU data obtained from 16S rRNA sequencing and
were processed and cleaned using the R package caret (v6.0.73).
Optimization of the mtry hyper-parameter involved making 100 different
80/20 (train/test) splits of the data where the same proportion was
present within both the whole data set and the 80/20 split. For each of
the different splits, 20 repeated 10-fold cross validation was performed
on the 80\% component to optimize the mtry hyper-parameter by maximizing
the AUC (Area Under the Curve of the Receiver Operator Characteristic).
The resulting model was then tested on the hold out data obtained from
the 20\% component. For all pre versus post-treatment models the
optimized mtry was 2 and for all normal versus diagnosis models the
optimized mtry was 2. The hyper-parameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
was created with the Random Forest model {[}33{]}.

@@

\textbf{\emph{Statistical Analysis.}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}34{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall
differences in the median relative abundance of each OTU between pre and
post-treatment samples utilized a paired Wilcoxon ranked sum test. Where
multiple comparison testing was appropriate, a Benjamini-Hochberg (BH)
correction was applied {[}35{]} and a corrected P-value of less than
0.05 was considered significant. The P-values reported are those that
were BH corrected. Model rank importance was determined by obtaining the
median MDA from the 100, 20 repeated 10-fold cross validation and then
ranking from largest to smallest MDA.

@@

\textbf{Figure 1: General differences between adenoma, advanced adenoma,
and carcinoma groups after treatment.} A) Thetayc distance from pre
versus post sample within each individual. A significant difference was
found between the adenoma and carcinoma group for thetayc (P-value =
5.36e-05). Solid black points represent the median value for each
diagnosis group. B) NMDS of the pre and post-treatment samples for the
adenoma group. C) NMDS of the pre and post-treatment samples for the
advanced adenoma group. D) NMDS of the pre and post-treatment samples
for the carcinoma group.

\textbf{Figure 2: The 10 OTUs used to classify treatment for adenoma,
advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
OTUs. C) Carcinoma OTUs.

@@

\begin{longtable}[]{@{}lccc@{}}
\toprule
& Adenoma & Advanced Adenoma & Carcinoma\tabularnewline
\midrule
\endhead
n & 22 & 19 & 26\tabularnewline
Age (Mean ± SD) & 61.68 ± 7.2 & 63.11 ± 10.9 & 61.65 ±
12.9\tabularnewline
Sex (\%F) & 36.36 & 36.84 & 42.31\tabularnewline
BMI (Mean ± SD) & 26.86 ± 3.9 & 25.81 ± 4.7 & 28.63 ± 7.2\tabularnewline
Caucasian (\%) & 95.45 & 84.21 & 96.15\tabularnewline
Days Between Colonoscopy (Mean ± SD) & 255.41 ± 42 & 250.16 ± 41 &
350.85 ± 102\tabularnewline
\bottomrule
\end{longtable}

@@

\hypertarget{ref-obrien_impact_2013}{}
27. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.

\hypertarget{ref-kozich_development_2013}{}
28. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq Illumina sequencing
platform. Applied and Environmental Microbiology. 2013;79:5112--20.

>>>>

\section{Normalization of the microbiota in patients after treatment for
colonic
lesions}\label{normalization-of-the-microbiota-in-patients-after-treatment-for-colonic-lesions}

@@

\textbf{Results.} There were small changes to the bacterial community
associated with adenoma or advanced adenoma groups and large changes
associated with the carcinoma group. The communities from patients with
carcinomas changed significantly more than those with adenoma following
treatment (P-value \textless{} 0.001). There was no significant change
in the microbiota between patients with adenoma and advanced adenoma, or
between patients with advanced adenoma and carcinoma (P-value
\textgreater{} 0.05). Although treatment was associated with
intrapersonal changes, the change in the abundance of individual OTUs to
treatment was not consistent within diagnosis groups (P-value
\textgreater{} 0.05). Because the distribution of OTUs across patients
and diagnosis groups was irregular, we used the Random Forest machine
learning algorithm to try and identify groups of OTUs that would allow
us to successfully distinguish between pre and post-treatment samples
for each of the diagnosis groups. Although the three models could
successfully differentiate between the pre and post-treatment samples,
there was little overlap between the OTUs that were indicative of
treatment and randomization of the diagnosis labels recapitulated the
actual models AUCs. Next, we used a larger cohort that contained
individuals with normal colons and those with adenomas, advanced
adenomas, and carcinomas to determine whether individuals who underwent
treatment were more likely to have OTUs associated with normal colons.
We again built Random Forest models and measured the change in the
positive probability of having one of the three diagnoses to assess
whether these changes were still similar to their original diagnosis or
not. Only patients who had carcinomas had a change that was towards a
microbial milieu that resembles the normal colon (P-value \textless{}
0.05). Finally, we tested whether the type of treatment impacted the
microbiota of those diagnosed with carcinomas and were unable to detect
any significant differences in characteristics of these communities
between individuals treated with surgery alone and those treated with
chemotherapy or chemotherapy and radiation (P-value \textgreater{}
0.05).

\textbf{Conclusions.} Although it was difficult to identify significant
differences between pre and post treatment samples for adenoma and
advanced adenoma, this was not the case for carcinomas. Not only were
there large changes in pre versus post treatment samples for those with
carcinoma but also these changes were towards a more normal microbiota
composition.

@@

\textbf{\emph{Treatment for colonic lesions alters the bacterial
community structure}} Within our 67-person cohort we tested whether the
microbiota of patients with adenoma (N = 22), advanced adenoma (N = 19),
or carcinoma (N = 26) had any broad differences between pre and
post-treatment samples {[}Table 1{]}. None of the individuals in this
study had any recorded antibiotic use before treatment of their
respective lesion {[}Table 1{]}. The structure of the microbial
communities of the pre and post-treatment samples differed, as measured
by the \(\theta\)\textsubscript{YC} beta diversity metric {[}Figure
1A{]}. We found that the communities obtained pre and post-treament
among the patients with carcinomas changed significantly more than those
patients with adenoma (P-value \textless{} 0.001). There were no
significant differences in the amount of change observed between the
patients with adenoma and advanced adenoma or between the patients with
advanced adenoma and carcinoma (P-value \textgreater{} 0.05). When
analyzing only those who underwent surgery as a treatment these trends
persisted (P-value \textgreater{} 0.05) {[}Figure S1A{]}. There was also
no difference in the change of \(\theta\)\textsubscript{YC} between pre
and post treatment samples between those who had surgery versus those
who did not for both the adenoma and advanced adenoma group (P-value
\textgreater{} 0.05) {[}Figure S1B-C{]}. Next, we tested whether there
was a consistent direction in the change in the community structure
between the pre and post-treatment samples for each of the diagnosis
groups {[}Figure 1B-D{]}. We only observed a consistent shift in
community structure for the patients with carcinoma when using a
PERMANOVA test (adenoma P-value = 0.999, advanced adenoma P-value =
0.945, and carcinoma P-value = 0.005). Finally, we measured the number
of observed OTUs, Shannon evenness, and Shannon diversity in the pre and
post-treatment samples and did not observe a significant change for any
of the diagnosis groups (P-value \textgreater{} 0.05) {[}Table S1{]}.

\textbf{\emph{The treatment of lesions are not consistent across
diagnosis groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups (P-value
\textgreater{} 0.05). It is likely that high inter-individual variation
and the irregular distribution of OTUs across individuals limited the
statistical power of the test. We attempted to overcome these problems
by using Random Forest models to identify collections of OTUs that would
allow us to differentiate between pre and post-treatment samples from
each of the diagnosis groups. To try and limit the likelihood that the
models would overfit the data because of the relatively small number of
subjects in each group, we restricted our models to only incorporate 10
OTUs. Despite this restriction, the models performed well (adenoma AUC
range = 0.69 - 0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma
AUC range = 0.82 - 0.98). However, upon randomizing the pre and post
treatment labels we found that Random Forest could work equally well
(adenoma AUC range = 0.73 - 0.98, advanced adenoma AUC range = 0.70 -
0.97, carcinoma AUC range = 0.89 - 0.99). Interestingly, the 10 OTUs
that were used for each model had little overlap with each other
{[}Figure 2{]}. Although treatment had an impact on the overall
community structure; the effect of treatment was not consistent across
patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma more
closely resemble those of a normal colon.}} Next, we determined whether
treatment changed the microbiota in a way that the post-treatment
communities resembled that of patients with normal colons. To test this,
we used an expanded cohort of 423 individuals that were diagnosed under
the same protocol as having normal colons or colons with adenoma,
advanced adenoma, or carcinoma {[}Table 2{]}. We then constructed Random
Forest models to classify the study samples, with the 3 diagnosis groups
(adenoma, advanced adenoma, or carcinoma), or having a normal colon. The
models performed well (adenoma AUC range =0.62 - 0.72, advanced adenoma
AUC range = 0.68 - 0.77, carcinoma AUC range = 0.84 - 0.90; Figure S2).
The OTUs that were incorporated into the adenoma and advanced adenoma
models largely overlapped and those OTUs that were used to classify the
carcinoma samples were largely distinct from those of the other two
models {[}Figure 3A{]}. Among the OTUs that were shared across the three
models were those populations generally considered beneficial to their
host (e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
\emph{Roseburia}) {[}Figures 3B{]}. Although many of these OTUs were
also included in the model differentiating between patients with normal
colons and those with carcinoma, this model also included OTUs
affiliated with populations that have previously been associated with
carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
\emph{Parvimonas}) {[}24--26{]} {[}Figure S3{]} with some individuals
showing are marked decrease in relative abundance {[}Figure S4{]}.
Finally, we applied these three models to the pre and post-treatment
samples for each diagnosis group and quantified the change in the
positive probability of the model. A decrease in the positive
probability would indicate that the microbiota more closely resembled
that of a patient with a normal colon. There was no significant change
in the positive probability for the adenoma or advanced adenoma groups
(P-value \textgreater{} 0.05) {[}Figure 4{]}. The positive probability
for the pre and post-treatment samples from patients diagnosed with
carcinoma significantly decreased with treatment, suggesting a shift
toward a normal microbiota for most individuals (P-value = 0.0156).
Only, 6 of the 26 patients (23.08\%) who were diagnosed with a carcinoma
had a higher positive probability after treatment; one of those was
re-diagnosed with carcinoma on the follow up visit. These results
indicate that, although there were changes in the microbiota associated
with treatment, those experienced by patients with carcinoma after
treatment yielded gut bacterial communities of greater similarity to
that of a normal colon.

\textbf{\emph{Difficult to identify effects of specific treatments on
the change in the microbiota.}} The type of treatment that the patients
received varied across diagnosis groups. Those with adenomas and
advanced adenomas received surgical resection (adenoma, N=4; advanced
adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=18;
advanced adenoma, N=15) and those with carcinomas received surgical
resection (N=12), surgical resection with chemotherapy (N=9), and
surgical resection with chemotherapy and radiation (N=5). Regardless of
treatment used there was no significant difference in the effect of
these treatments on the number of observed OTUs, Shannon diversity, or
Shannon evenness (P-value \textgreater{} 0.05). Furthermore, there was
no significant difference in the effect of the treatments on the amount
of change in the community structure (P-value = 0.532). Finally, the
change in the positive probability was also not significantly different
between any treatment group (P-value = 0.532). Due to the relatively
small number of samples in each treatment group, it was difficult to
make a definitive statement regarding the specific type of treatment on
the amount of change in the structure of the microbiota.

@@

Our study focused on comparing the microbiota of patients diagnosed with
adenoma, advanced adenoma, and carcinoma before and after treatment. For
all three groups of patients, we observed changes in their microbiota.
Some of these changes, specifically for adenoma, may be due to normal
temporal variation but those with advanced adenoma and CRC clearly had
large microbiota changes that would be greater than what could be
expected from this type of variation {[}27{]}. After treatment, the
microbiota of patients with carcinoma changed significantly more than
the other groups. This change resulted in communities that more closely
resembled those of patients with a normal colon. This may suggest that
treatment for carcinoma is not only successful for removing the
carcinoma but also at reducing the associated bacterial communities.
Understanding the effect of treatment on the microbiota of those
diagnosed with carcinomas may have important implications for reducing
disease recurrence. It is intriguing that it may be possible to use
microbiome-based biomarkers to not only predict the presence of lesions
but to also assess the risk of recurrence.

Patients diagnosed with adenoma and advanced adenoma, however, did not
experience a shift towards a community structure that resembled those
with normal colons. This may be due to the fundamental differences
between the features of adenomas and advanced adenomas and carcinoma.
Specifically, carcinomas may create an inflammatory milieu that would
impact the structure of the community and removal of that stimulus would
alter said structure. It is possible that the difference between the
microbiota of patients with adenoma and advanced adenoma and those with
normal colons is subtle. This is supported by the reduced ability of our
models to correctly classify patients with adenomas and advanced
adenomas relative to those diagnosed with carcinomas {[}Figure S2{]}.
Given the irregular distribution of microbiota across patients in the
different diagnosis groups, it is possible that we lacked the
statistical power to adequately characterize the change in the
communities following treatment.

@@

This study expands upon existing research that has established a role
for the microbiota in tumorigenesis and that demonstrated the utility of
microbiome-based biomarkers to predict the presence of colonic lesions.
Although in the adenoma and advanced adenoma groups we cannot make a
definitive statement regarding the effect of a specific treatment on the
gut microbiota, due to the low number of samples, small effect sizes,
potential changes in diet after treatment, changing medical care for
chronic maladies, etc. There are still a number of exciting new
potential avenues to explore based on the carcinoma observations. One of
the most exciting of these future directions is the possibility that
markers within the microbiota could be used to potentially evaluate the
effect of treatment and predict recurrence for those diagnosed with
carcinoma. If such an approach is effective, it might be possible to
target the microbiota as part of adjuvant therapy, if the biomarkers
identified play a key role in the disease process. Our data provides
additional evidence on the importance of the microbiota in tumorigenesis
by addressing the recovery of the microbiota after treatment and opens
interesting avenues of research into how these changes may affect
recurrence.

@@

\textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
have been previously reported in Baxter, et al {[}24{]}. Briefly,
samples were stored on ice for at least 24h before freezing. Although we
cannot exclude that this sampling protocol may have impacted the gut
microbiota composition all samples were subjected to the same
methodology. Study exclusion involved those who had already undergone
surgery, radiation, or chemotherapy, had colorectal cancer before a
baseline fecal sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or familial adenomatous polyposis.
Samples used to build the models for prediction were collected either
prior to a colonoscopy or between one and two weeks after initial
colonoscopy. The bacterial community has been shown to normalize back to
a pre-colonoscopy community within this time period {[}28{]}. Our study
cohort consisted of 67 individuals with an initial sample as described
and a follow up sample obtained between 188 - 546 days after treatment
of lesion {[}Table 1{]}. Patients were diagnosed by colonoscopic
examination and histopathological review of any biopsies taken. Patients
were classified as having advanced adenoma if they had an adenoma
greater than 1 cm, more than three adenomas of any size, or an adenoma
with villous histology. This study was approved by the University of
Michigan Institutional Review Board. All study participants provided
informed consent and the study itself conformed to the guidelines set
out by the Helsinki Declaration.

\textbf{\emph{Treatment}} For this study treatment refers specifically
to the removal of a lesion or with or without chemotherapy and
radiation. The majority of patients undergoing treatment for adenoma or
advanced adenoma were not treated surigically {[}Table 1{]} but rather
via colonoscopy. All CRC individuals were treated with at least surgery
or a combination of surgery and chemotherapy or surgery, chermotherapy,
and radiation. The type of chemotherapy used for patients with CRC
varied markably. The types used included Oxaliplatin, Levicovorin,
Folfox, Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin.
These were used individually or in combination with others depending on
the patient {[}Table 1{]}. If an individual was treated with radiation
they were also always treated with chemotherapy. Radiation therapy
generally used 18 mV photons for treatment.

\textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
described by Kozich, et al. {[}29{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between four sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing.}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}29{]}. The general workflow using mothur
included merging paired-end reads into contigs, filtering for low
quality contigs, aligning to the SILVA database {[}30{]}, screening for
chimeras using UCHIME {[}31{]}, classifying with a naive Bayesian
classifier using the Ribosomal Database Project (RDP){[}32{]}, and
clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm
{[}33{]}. The number of sequences for each sample was rarefied to 10523
to minimize the impacts of uneven sampling.

\textbf{\emph{Model Building.}} The Random Forest {[}34{]} algorithm was
used to create the three models used to classify pre and post-treatment
samples by diagnosis (adenoma, advanced adenoma, or carcinoma). The
total number of individuals in the pre versus post-treatment models was
67 individuals. There were a total of 22 individuals in the pre versus
post-treatment adenoma model, 19 individuals in the pre versus
post-treatment advanced adenoma model, and 26 individuals in the pre
versus post-treatment carcinoma model {[}Table 1{]}.

Similarily, the Random Forest {[}34{]} algorithm was also used to create
the three models used to classify normal versus diagnosis. These samples
were obtained using the same methodology as described earlier in this
section. All samples used for this component of model training were from
pre-treatment samples. The total number of individuals in the normal
versus diagnosis models was 423 individuals {[}Table 2{]}. There were a
total of 239 individuals in the normal versus adenoma model, 262
individuals in the normal versus advanced adenoma model, and 266
individuals in the normal versus carcinoma model {[}Table 2{]}.

All models included only OTU data obtained from 16S rRNA sequencing and
were processed and cleaned using the R package caret (v6.0.76).
Optimization of the mtry hyper-parameter involved making 100 different
80/20 (train/test) splits of the data where the same proportion was
present within both the whole data set and the 80/20 split. For each of
the different splits, 20 repeated 10-fold cross validation was performed
on the 80\% component to optimize the mtry hyper-parameter by maximizing
the AUC (Area Under the Curve of the Receiver Operator Characteristic).
The resulting model was then tested on the hold out data obtained from
the 20\% component. For all pre versus post-treatment models the
optimized mtry was 2 and for all normal versus diagnosis models the
optimized mtry was 2. The hyper-parameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
was created with the Random Forest model {[}34{]}.

@@

\textbf{\emph{Statistical Analysis.}} The R software package (v3.4.1)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}35{]} in the vegan package
(v2.4.3). Comparisons between probabilities as well as overall
differences in the median relative abundance of each OTU between pre and
post-treatment samples utilized a paired Wilcoxon ranked sum test.
Comparisons between different treatment for lesions utilized a Kruskal
Wallis test. Where multiple comparison testing was appropriate, a
Benjamini-Hochberg (BH) correction was applied {[}36{]} and a corrected
P-value of less than 0.05 was considered significant. The P-values
reported are those that were BH corrected. Model rank importance was
determined by obtaining the median MDA from the 100, 20 repeated 10-fold
cross validation and then ranking from largest to smallest MDA.

@@

@@

\begin{longtable}[]{@{}lccc@{}}
\toprule
& Adenoma & Advanced Adenoma & Carcinoma\tabularnewline
\midrule
\endhead
n & 22 & 19 & 26\tabularnewline
Age (Mean ± SD) & 61.68 ± 7.2 & 63.11 ± 10.9 & 61.65 ±
12.9\tabularnewline
Sex (\%F) & 36.36 & 36.84 & 42.31\tabularnewline
BMI (Mean ± SD) & 26.86 ± 3.9 & 25.81 ± 4.7 & 28.63 ± 7.2\tabularnewline
Caucasian (\%) & 95.45 & 84.21 & 96.15\tabularnewline
Days Between Colonoscopy (Mean ± SD) & 255.41 ± 42 & 250.16 ± 41 &
350.85 ± 102\tabularnewline
Surgery Only & 4 & 4 & 12\tabularnewline
Surgery \& Chemotherapy & 0 & 0 & 9\tabularnewline
Surgery, Chemotherapy, \& Radiation & 0 & 0 & 5\tabularnewline
Pre-Treatment Antibiotic Use & 0 & 0 & 0\tabularnewline
\bottomrule
\end{longtable}

@@

\hypertarget{ref-schloss_dynamics_2014}{}
27. Schloss PD, Iverson KD, Petrosino JF, Schloss SJ. The dynamics of a
family's gut microbiota reveal variations on a theme. Microbiome.
2014;2:25.

\hypertarget{ref-obrien_impact_2013}{}
28. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
